---
title: "TP2: Statistiques inférentielles: Intervalles de confiance"
output: html_notebook
---



```{r}
# Constants
sigma <- 9
mu <- 100
n <- 36 # sample size
```


```{r}
# set the seed used by random generator numbers (useful for debug and reproducibility of the results
set.seed(123)
sample <- rnorm(n=n,mean=mu,sd=sigma)
# display histogram of the sample (Remark sample size is low)
hist(sample, main=paste("Histogram of a sample of size", n, "drawn from a normal law N(",mu,",", sigma^2,"))"),col="lightgreen",xlab="")
```
```{r}
#' Compute confidence interval associated with mean of a sample 
#' 
#' @param sample a numeric vector
#' @param std theoretical standard deviation .
#' @returns ci a numeric vector with the two bounds of the confidence interval
#' @examples
#' confidence_interval_mean(sample, 9, 0.05)
#' 
confidence_interval_mean <- function(sample, std, alpha){
  mean_sample <- mean(sample)
  width <- std/sqrt(length(sample))*qnorm(1-(alpha/2))
  ci <- c(mean_sample - width, mean_sample + width)
  return(ci)
}

#' Compute the width of confidence interval given its two bounds
#' 
#' @param ci confidence interval bounds, c(lower_bound,upper_bound)
#' @returns ,float
#' @examples
#' ci_width(confidence_interval)
#'
ci_width <- function(ci){
  width = ci[2] - ci[1]
  return(width)
}

#' Compute confidence interval associated with the mean estimate from a sample 
#' and display results in the terminal
#' 
#' @param sample numeric vector
#' @param sigma numeric vector
#' @param alpha float  
#' @returns sig
#' @examples
#' ci_width(confidence_interval)
#'
display_results <-function(sample,sigma, alpha=0.05){
ci <- confidence_interval_mean(sample, sigma, alpha)
cat("Confidence interval associated with the mean estimate of the sample at risk level alpha=",alpha, "is:\n","ci=[",ci,"]","\n")
# interval width 
ci_width_default <- ci_width(ci)
cat("Confidence interval width associated with the mean estimate of the sample at risk level alpha=",alpha, "is:\n",ci_width_default,"\n")
}
```


```{r}
for (alpha in c(0.05, 0.1)){
  display_results(sample, sigma, alpha=alpha)
}
```
On remarque que :

+ l'estimation ponctuelle de la moyenne est identique (l'échantillon n'a pas été regénéré)
+ la largeur de l'intervalle de confiance évolue en $u_{1-\frac{\alpha}{2}}\frac{\sigma}{\sqrt{n}}$ avec $u_{1-\frac{\alpha}{2}}$ le quantile d'ordre $1-\frac{\alpha}{2}$ de la loi $N(0,1)$.
+ à taille d'échantillon $n$ et écart type $\sigma$ fixé, la taille de l'intervalle de confiance associée à la moyenne diminue quand $\alpha$ augmente.

```{r}
is_mean_in_ci <-function(ci, theoretical_mean){

  lower_inclusion <- (theoretical_mean >= ci[1])
  upper_inclusion <- (theoretical_mean <= ci[2])
  inclusion_test <- (lower_inclusion && upper_inclusion)
  
return(inclusion_test) 
}

estimate_frequency_in <-function(sample_size, mean, sigma, n_repetition, alpha){
  n_in <- 0
  for (i in 1:n_repetition){
   # generate a new normal sample of size sample_size according to N(mean, sigma)
   sample <- rnorm(n=sample_size,mean=mean,sd=sigma)
   # compute confidence interval associated with mean estimate
   ci <- confidence_interval_mean(sample, sigma, alpha)
   # test if theoretical mean is in the estimated confidence interval of the mean
   test <- is_mean_in_ci(ci, mu)
   if (test == TRUE){
     n_in <- (n_in + 1)
   }
  }
  # compute frequency to which theoretical mean is in the confidence interval of the mean
  f_mean_in = n_in/n_repetition
  return(f_mean_in) 
}


```

```{r}
n_repetition <- 1000
for (alpha in c(0.05,0.1)){
  f_mean_in <- estimate_frequency_in(n,mu,sigma, n_repetition, alpha)
  cat("Pourcentage du nombre d'occurence de la vraie moyenne dans l'intervalle de confiance pour un risque   alpha=",alpha, ":",f_mean_in, "\n")
}

```


## La fonction t.test
```{r}
# Lets take the exact same sample
set.seed(123)
n <- 36
sample <- rnorm(n=n,mean=mu,sd=sigma)
ci_095_unknown_var <- t.test(sample)$conf.int
ci_width_095_unknown_var <- ci_width(ci_095_unknown_var)
ci_090_unknown_var <- t.test(sample,conf.level = 0.90)$conf.int
ci_width_090_unknown_var <- ci_width(ci_090_unknown_var)


cat("Width of the confidence interval associated with the mean at risk level alpha=0.05 estimated using t.test", ci_width_095_unknown_var,"\n")

cat("Width of the confidence interval associated with the mean at risk level alpha=0.1 estimated using t.test", ci_width_090_unknown_var,"\n")
```

La largeur de l'intervalle de confiance associé à la moyenne quand l'ecart type est connu est donnée par
$2u_{1-\frac{\alpha}{2}}\frac{\sigma}{n}$ à mettre en regard avec $2t_{1-\frac{\alpha}{2}}(n-1)\frac{s^*}{n}$.


$t_{1-\frac{\alpha}{2}}(n-1)> u_{1-\frac{\alpha}{2}}$ l'intervalle de confiance associé à l'estimation de la moyenne sans connaitre la variance est donc plus large et ce particulièrement lorsque $n$ est petit. Lorsque la taille d'échantillon $n$ augmente la distribution de Student tend vers une loi normale, les largeurs des intervalles de confiance deviennent équivalentes. 
